Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-23 analysis/paper/paper.Rmd
additions to Intro

***Paragraph 2***

One potentially important, but perhaps under-appreciated source of mortality on Chinook salmon is predation on emigrating juveniles by piscivorous fishes, including both native and non-native species. As an example,...

  - At this point, I'm thinking leave avian predation out of the equation, but instead include in Discussion
  - and keep focus on the freshwater environment
* What is known about downriver (Columbia/Snake) piscivorous fish populations and potential predation on juvenile Chinook salmon?
  - literature review of northern pikeminnow, smallmouth bass, other? predation on juvenile Chinook salmon, particulary from the Snake and Columbia rivers
  - still stay away from avian predation

Northern Pikeminnow are estimated to consume 8% (16.4 million) of the estimated 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996]. Dams and reservoirs in the Columbia River are associated with high rates of predation on salmonids [@Petersen1994; @Ward1995]. Because migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], migrating smolts may become more vulnerable to predation in areas with slower velocities. Most of the research has been targeted around dams and reservoirs, but there are areas functioning similar to a reservoir in the ruvers upstream. One of these areas is the Deadwater Slough in the Salmon River. The Deadwater Slough is a 25.3-acre area that is unnaturally slow and deep within the Salmon River and resembles a small reservoir (Figure \@ref(fig:map)). This unnatural area has little to no cover, higher depth, and fine substrate which favor piscivorous fish predators (e.g., northern pikeminnow *Ptychocheilus oregonensis*, smallmouth bass *Micropterus dolomieu*) [@Watkins2015]. @Axel2015 demonstrated decreased rates of emigration and apparent survival for Sockeye Salmon *Oncorhynchus nerka* from Redfish Lake emigrating during the spring. Further, recent winter telemetry studies have indicated decreased transition probabilities (approximately 10% less than surrounding reaches) of juvenile Chinook salmon through Deadwater Slough during fall and winter months [@Ackerman2018; @Porter2019].

>>>>

One potentially important, but perhaps under-appreciated source of mortality for Chinook Salmon is predation on emigrating juveniles by piscivorous fishes, including both native and non-native species. Dams and reservoirs in the Columbia River basin are the primary locations associated with high rates of predation on salmonids [@Petersen1994; @Ward1995]. There are generally two maechanisms that explain these high predation zones. First, migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces created by dams on the Columbia and Snake rivers have created favorable slow-water habitat for several species of piscivorous fish known to consume juvenile Chinook Salmon including Walleye *Sander vitreus*, Smallmouth Bass *Micropterus dolomieu*, and Northern Pikeminnow *Ptychocheilus oregonensis*. In the Lower Granite Reservoir on the Snake River, juvenile Chinook Salmon are common prey items of non-native Smallmouth Bass during some seasons [@Erhardt2018]. Walleye also prefer slow water habitats and are considered a potentially important source for Juvenile Chinook Salmon mortality [@NOAA2020], particularly as their abundance has significantly increased in Snake River dam facilities between 2008 and 2020.

In addition to non-native species, Northern Pikeminnow are a native, piscivorous fish preferring slow water habitats. Consequently, they too have benefited from dams on the Columbia River and become abundance predators on salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the estimated 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996]. Predation by piscivorous fish may have been a factor in the below-average survival of wild Chinook Salmon in 2020 [@NOAA2020], however there is a dearth of current data on piscivorous predator populations in the Columbia River.

Habitat containing slow-velocities and other attributes that support piscivorous predators are not limited to dams and reservoirs. One such area is the Deadwater Slough in the Salmon River, which occupies a 25.3-acre reach of unnaturally slow and deep water (Figure \@ref(fig:map)). The Deadwater Slough contains little to no cover, increased depth, and fine substrate which favor piscivorous fish predators (e.g., Northern Pikeminnow, Smallmouth Bass) [@Watkins2015]. @Axel2015 demonstrated decreased rates of emigration and apparent survival for Sockeye Salmon *Oncorhynchus nerka* from Redfish Lake emigrating during the spring. Further, recent winter telemetry studies have indicated decreased transition probabilities (approximately 10% less than surrounding reaches) of juvenile Chinook salmon through Deadwater Slough during fall and winter months [@Ackerman2018; @Porter2019].
